---
title: "Snapshot Conversion: Binary Wrapper and Pure-R Fallbacks"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Snapshot Conversion: Binary Wrapper and Pure-R Fallbacks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  cache: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

`openalexPro` provides two implementations for every snapshot processing task:

| Task | Binary wrapper (preferred) | Pure-R fallback |
|------|---------------------------|-----------------|
| Convert JSON → Parquet | `snapshot_to_parquet()` | `snapshot_to_parquet_R()` |
| Build ID lookup index | `build_corpus_index()` | `build_corpus_index_R()` |
| Extract records by ID | `lookup_by_id()` | `lookup_by_id_R()` |

**Binary wrappers** delegate to the `openalex-snapshot` Rust CLI, which is
faster, includes built-in verification and repair, and handles very large
datasets with lower memory overhead.

**Pure-R fallbacks** (`*_R()` variants) implement the same logic entirely in
R using DuckDB, with no external binary dependency. They are suitable when
installing the binary is not possible, or for understanding implementation
internals.

Both sets of functions are exported and fully documented.

# Binary Wrappers

## Prerequisite — the `openalex-snapshot` Binary

Download from [GitHub releases](https://github.com/rkrug/openalex-snapshot/releases)
or build from source:

```bash
cargo build --release
```

Once installed, place the binary on your PATH, or provide its location via:

```{r}
# Option 1: set for the session
options(openalexPro.oas_bin = "/path/to/openalex-snapshot")

# Option 2: pass directly to each function call
snapshot_to_parquet(root_dir = ".", oas_bin = "/path/to/openalex-snapshot")
```

## Root Directory Layout

All three binary wrappers use a single `root_dir` parameter. The binary
derives all paths from it:

```
<root_dir>/
  openalex-snapshot/          # raw JSON .gz files (input)
  parquet/                    # converted parquet files (output)
  parquet/<dataset>_id_idx.parquet  # lookup index
  .openalex-snapshot_metadata/      # metadata and checksums
```

If `root_dir = "."` (the default for the Makefile), the working directory is
the root.

## `snapshot_to_parquet()` — Convert JSON to Parquet

```{r}
library(openalexPro)

# Convert all datasets (works, authors, institutions, ...)
snapshot_to_parquet(
  root_dir  = "/Volumes/openalex",
  workers   = 4,
  profile   = "balanced",     # "safe" | "balanced" | "fast"
  max_memory_mb = 15000
)

# Convert a specific dataset only
snapshot_to_parquet(
  root_dir  = "/Volumes/openalex",
  data_sets = "works",
  workers   = 4
)
```

`data_sets = NULL` (default) processes all datasets. When a character vector
is given, each dataset is passed to the binary in a separate call (the binary
accepts one `--dataset` per invocation).

The function returns `invisible(root_dir)`.

## `build_corpus_index()` — Build ID Lookup Indexes

After conversion, build an ID index for fast lookups:

```{r}
build_corpus_index(
  root_dir  = "/Volumes/openalex",
  data_sets = NULL,          # NULL = all datasets
  workers   = 8,
  profile   = "balanced",
  overwrite = FALSE          # TRUE = rebuild even if index exists
)
```

This creates `<root_dir>/parquet/<dataset>_id_idx.parquet` for each dataset.
Each index maps OpenAlex IDs to their physical location (file and row number),
enabling fast record retrieval without scanning the full corpus.

## `lookup_by_id()` — Extract Records by OpenAlex ID

```{r}
# Extract specific works into a project directory
lookup_by_id(
  root_dir    = "/Volumes/openalex",
  ids         = c("W2741809807", "W2100837269", "W1234567890"),
  project_dir = "my_project",
  data_sets   = "works"       # NULL = search all datasets
)
# Extracted parquet files are written to my_project/snapshot_extract/
```

IDs can be in short form (`"W2741809807"`) or long form
(`"https://openalex.org/W2741809807"`). Both are accepted and normalised
automatically.

The `project_dir` convention is consistent with `pro_fetch()` and
`pro_request()`: the output directory identifies a project, and results
accumulate inside it.

The function returns `invisible(project_dir)`.

## Performance Profiles

The `profile` argument controls the binary's memory/speed trade-off:

| Profile | Memory use | Speed | When to use |
|---------|-----------|-------|-------------|
| `"safe"` | Low | Slow | Shared machines, limited RAM |
| `"balanced"` | Medium | Medium | Default for most workloads |
| `"fast"` | High | Fast | Dedicated machines with ≥ 32 GB RAM |

---

# Pure-R Fallbacks

The `*_R()` functions contain the original R + DuckDB implementations,
preserved for environments where the binary is unavailable. They are
slower than the binary wrappers but have no external dependencies beyond
the `duckdb` and `arrow` R packages.

## `snapshot_to_parquet_R()` — Convert JSON to Parquet (R)

```{r}
snapshot_to_parquet_R(
  snapshot_dir   = "/Volumes/openalex/openalex-snapshot",
  parquet_dir    = "/Volumes/openalex/parquet",
  data_sets      = NULL,       # NULL = all; or e.g. c("works", "authors")
  sample_size    = 100,        # number of .gz files to sample for schema inference
  memory_limit   = "15GB",     # per-DuckDB-connection limit
  temp_directory = "/tmp",     # DuckDB spill-to-disk location
  workers        = 4           # parallel future workers
)
```

Note that `snapshot_to_parquet_R()` takes **separate** `snapshot_dir` and
`parquet_dir` parameters, unlike the binary wrapper which derives both from
`root_dir`.

### Two-Stage Pipeline

**Stage 1 — Schema inference** (`infer_json_schema()`): runs `DESCRIBE SELECT *
FROM read_json_auto(file)` on a random sample of `.gz` files, then merges the
per-file schemas into a single widened schema using type-widening rules
(complex types beat simple types; widest numeric wins; fallback to `VARCHAR`).
Results are cached to `.schema_cache/` to support resume.

**Stage 2 — Per-file conversion**: each `.gz` file is converted to `.parquet`
using the unified schema from Stage 1. One DuckDB connection per worker; already-
converted files are skipped (resume support).

### Schema Cache

Schema inference is the slowest sequential step. A two-level cache avoids
repeating it:

```
parquet/works/.schema_cache/
  unified_schema.csv        ← loaded first; if found, no DuckDB queries run
  2024-01-15_part_000.csv   ← per-file cache; enables mid-run resume
  2024-01-15_part_001.csv
  ...
```

**To force re-inference**, delete `unified_schema.csv`:

```bash
rm parquet/works/.schema_cache/unified_schema.csv
```

**To pre-populate the cache** (e.g. after downloading an updated snapshot with
the same schema), copy the file from a previous successful run:

```bash
mkdir -p parquet/works/.schema_cache
cp ~/schemas/works_schema.csv parquet/works/.schema_cache/unified_schema.csv
```

### Special Case — `works` Dataset

The `works` dataset has two workarounds applied automatically by
`snapshot_to_parquet_R()`:

1. **Large object size**: `maximum_object_size=1000000000` (1 GB) is passed to
   DuckDB because individual work records with long author lists can exceed the
   default limit.

2. **`abstract_inverted_index` stored as VARCHAR**: OpenAlex encodes abstracts
   as an inverted index — a JSON object whose keys are words. DuckDB
   case-folds struct field names, causing a collision when the same word appears
   with different capitalisation (e.g. `"as"` and `"As"`). Storing the column
   as `VARCHAR` (raw JSON text) avoids the collision. To decode:

```{r}
library(arrow)
library(jsonlite)

works <- open_dataset("parquet/works")
row   <- works |> filter(id == "W2741809807") |> collect()

aii   <- fromJSON(row$abstract_inverted_index[1])
# aii is a named list: list("The" = 0, "quick" = 1, "as" = 3, "As" = 7)
```

## `build_corpus_index_R()` — Build ID Index (R)

```{r}
# Build index for one dataset
build_corpus_index_R(
  corpus_dir   = "/Volumes/openalex/parquet/works",
  memory_limit = "15GB",
  workers      = 4
)
# Creates /Volumes/openalex/parquet/works_id_idx.parquet
```

Note that `build_corpus_index_R()` takes a `corpus_dir` pointing to a
single dataset directory (e.g. `parquet/works`), whereas the binary wrapper
takes `root_dir` and an optional `data_sets` argument.

### How the Index Works

The index contains four columns:

| Column | Description |
|--------|-------------|
| `id` | Full OpenAlex URL (e.g. `https://openalex.org/W2741809807`) |
| `id_block` | `floor(numeric_id / 10000)` — used for block-based partitioning |
| `parquet_file` | Relative path to the source parquet file |
| `file_row_number` | 0-indexed row within that file |

The `id_block` grouping allows the lookup function to filter the index to just
the relevant partition before reading source files, avoiding a full index scan.

### Two-Stage Index Build

**Stage 1**: Each parquet file is indexed individually (one DuckDB connection
per worker), writing a temporary per-file parquet to a staging directory.

**Stage 2**: The staging directory is merged into a single `*_id_idx.parquet`
file using a sequential DuckDB connection. The staging directory is removed.

## `lookup_by_id_R()` — Extract Records by ID (R)

```{r}
# Returns a data frame (in-memory mode)
works <- lookup_by_id_R(
  index_file = "/Volumes/openalex/parquet/works_id_idx.parquet",
  ids        = c("W2741809807", "W2100837269")
)

# Writes to parquet files (out-of-memory mode, for millions of IDs)
lookup_by_id_R(
  index_file = "/Volumes/openalex/parquet/works_id_idx.parquet",
  ids        = large_id_vector,
  output     = "filtered_works",  # directory must not already exist
  workers    = 3
)
```

Note that `lookup_by_id_R()` takes `index_file` (path to the
`*_id_idx.parquet` file) and returns either a data frame or writes to `output`.
The binary wrapper uses `root_dir` + `project_dir` instead.

### How Lookup Works

1. IDs are normalised to long form (`https://openalex.org/...`).
2. The index is filtered with `dplyr::filter(id %in% ids)` to find the source
   file and row number for each match.
3. Results are grouped by source parquet file (`parquet_file`).
4. For each source file, a DuckDB query retrieves exactly the rows needed:
   ```sql
   SELECT * FROM read_parquet('file.parquet', file_row_number = true)
   WHERE file_row_number IN (42, 107, 3891)
   ```
5. In `output` mode, each chunk is written directly to a parquet file without
   loading into R memory.

---

# Comparison Summary

| Feature | Binary wrappers | Pure-R fallbacks (`*_R()`) |
|---------|----------------|--------------------------|
| External dependency | `openalex-snapshot` binary | None (duckdb + arrow only) |
| Speed | Faster | Slower |
| Verification & repair | Built in | Not available |
| `root_dir` convention | Yes | No (separate `snapshot_dir`/`parquet_dir`/`corpus_dir`) |
| Schema inference | Handled by binary | `infer_json_schema()` with caching |
| `lookup_by_id` output | `project_dir` folder | data frame or `output` directory |
| Resume support | Yes | Yes (per-file skip) |
| Parallel workers | Yes | Yes (`future_lapply`) |

## When to Use Which

Use the **binary wrappers** when:

- The binary is available on your PATH (or you can install it)
- You need maximum throughput on large datasets
- You want built-in verification after conversion

Use the **pure-R fallbacks** when:

- You cannot install the binary (e.g. locked-down institutional systems)
- You need to inspect or customise the schema inference step
- You are building workflows that must work without any system-level installs
