---
title: "Compatibility with `openalexR`"
format: 
  html:
    toc: true
    embed-resources: true
params:
  search_term: "biodiversity AND conservation AND IPBES"
  identifier: 
    - "W3045921891"
    - "W3046863325"
  verbose: true
execute: 
  cache: false
  eval: true
---

```{r}
#| include: false
#| label: knitr_setup
try(
  params <- rmarkdown::yaml_front_matter("inst/compatibility.qmd")$params,
  silent = TRUE
)
```


```{r}
#| label: setup
library(openalexPro)
library(openalexR)
library(dplyr)
library(tictoc)
```

# Compatibility between `openalexR` and `openalexPro`
One of the aims of this package is compatibility between `openalexPro` and `openalexR`. 

In this report, several searches under different conditions (filters, grouping, snowball searches) will be done and the results compared. As `openalexPro` is agnostic to the underlying structure of the returned datsets, structural equivalence is not always expected or possible. 

Thew search scenarios we will compare below are:
- [TODO] count of search
- search
  - abstract and title search
  - [TODO] full text search
  - [TODO] with filter
  - [TODO] return of selected fields
- snowball search (this is only done if the package [`openalexSnowball`](https://rkrug.r-universe.dev/openalexPro) is also installed )

As `openalexR` is fully memory based, whereas `openalexPro` fully file based, doifferent approaches need to be followed which will be discussed in the sections. 

# Basic title_and_abstract Search

## `openalexR`

First we fetch the works using `openalexR::oa_fetch()`:

```{r}
#| label: openalexR_fetch
tic()
fn <- file.path("openalexR", "works.rds")

dir.create(dirname(fn), recursive = TRUE, showWarnings = FALSE)
fetchR <- openalexR::oa_fetch(
  entity = "works",
  output = "tibble",
  title_and_abstract.search = params$search_term,
) |>
  dplyr::arrange(id)

saveRDS(
  fetchR,
  file = fn
)
toc()
```


## `openalexPro`

Now we will downloiad the results using `openalexPro::pro_fetch()`

```{r}
#| label: openalexPro_fetch
tic()
fetchPro <- pro_query(
  title_and_abstract.search = params$search_term,
  entity = "works"
) |>
  pro_fetch(
    project_folder = "openalexPro",
    overwrite = TRUE,
    verbose = params$verbose
  )
toc()
```



# Snowball Search

## `openalexR`

First we fetch the works using `openalexR::oa_snowball()`:

```{r}
#| label: openalexR_oa_snowball
tic()
fn <- file.path("openalexR", "snowball.rds")

dir.create(dirname(fn), recursive = TRUE, showWarnings = FALSE)

snowballR <- openalexR::oa_snowball(
  identifier = params$identifier,
  id_type = "original",
  verbose = params$verbose
)
saveRDS(snowballR, fn)
toc()
```

## `openalexPro`

Now we will do the same using `openalexSnowball::pro_snowball()`.

```{r}
#| label: openalexPro_pro_snowball
tic()
snowballPro <- file.path("openalexPro", "snowball")

snowballPro <- openalexSnowball::pro_snowball(
  identifier = params$identifier,
  output = snowballPro,
  verbose = params$verbose
)
toc()
```

# Comparison Search Term Search

To compare the search term results, we have to read these

```{r}
#| label: read_search_results
fetchR <- readRDS(file.path("openalexR", "works.rds"))
fetchPro <- openalexPro::read_corpus(
  file.path(("openalexPro"), "parquet"),
  return_data = TRUE
)
```

## Works retrieved (`id`s)

Works retrieved using `openalexR::oa_fetch()` and not in the `openalexPro` result:

```{r}
setdiff(
  fetchR$id,
  fetchPro$id
)
```


Works retrieved using the `openalexPro` pipeline and not in `openalexR::oa_fetch()`:

```{r}
setdiff(
  fetchPro$id,
  fetchR$id
)
```

If both are empty, the same works are retrieved.

## Names of Columns

Let's compare the names. The following columns are in both `fetchR` and `fetchPro`:
```{r}
#| label: columns_in_both
intersect(
  names(fetchR),
  names(fetchPro)
)
```

The following are missing from `fetchPro`:
```{r}
#| label: columns not_in_fetchPro_comp
#| 
setdiff(
  names(fetchR),
  names(fetchPro)
)
```

and these are missing from `fetchR`:
```{r}
#| label: columns not_in_fetchR
#| 
setdiff(
  names(fetchPro),
  names(fetchR)
)
```

These differences are based on normalisation which `openalexR` os doing, while `openalexPro` is agnostic to the fields retrieved. See the documentation of `openalexR` for details.

# Comparison Snowball Search

To compare the snowball searches the results have to be read into memory.

```{r}
#| label: read_snowball_results
#| 
snowballR <- readRDS(file.path("openalexR", "snowball.rds"))
snowballPro <- openalexSnowball::read_snowball(
  file.path(("openalexPro"), "snowball"),
  return_data = TRUE
)
```

## Nodes - works retrieved (`id`s)

First we check if the same papers are retrieved:.

The following works are in `openalexR::oa_snowball()` but not in `openalexSnowball::pro_snowball()`:

```{r}
setdiff(
  snowballR$nodes$id,
  snowballPro$nodes$id
)
```

The following works are in `openalexR::oa_snowball()` but not in `openalexSnowball::pro_snowball()`:

```{r}
setdiff(
  snowballPro$id,
  snowballR$nodes$id
)
```

Both should be empty!

The fields in the nodes follow the same pattern as the default fetch.

## Edges


### Edges Identified

The following edges are in `openalexR::oa_snowball()` but not in `openalexSnowball::pro_snowball()`:

```{r}
setdiff(
  snowballR$edges |>
    dplyr::arrange(
      from,
      to
    ),
  snowballPro$edges |>
    dplyr::filter(edge_type == "core") |>
    dplyr::select(
      from,
      to
    )
)
```
  
The following works are in `openalexR::oa_snowball()` but not in `openalexSnowball::pro_snowball()`:

```{r}
setdiff(
  snowballPro$edges |>
    dplyr::filter(edge_type == "core") |>
    dplyr::select(
      from,
      to
    ),
  snowballR$edges |>
    dplyr::arrange(
      from,
      to
    )
)
```
